---
title: "Covid-19 data visualization: Canada"
author: "Kyle Deng"
date: "Last updated at `r format(Sys.time(), '%d %B, %Y - %H:%M',tz='America/Vancouver')`"
output: 
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
                      autodep = TRUE, fig.retina = 2,	dpi = 150, 
                      fig.width = 7.5, fig.height = 5)

options(scipen = 999)
library(tidyverse)
library(plotly)
library(zoo)
library(lubridate)
library(sf)
library(rnaturalearth)

CA <- read_csv("https://covid.ourworldindata.org/data/owid-covid-data.csv") %>%
  filter(location == "Canada") %>%
  drop_na(new_cases_smoothed) %>%
  select(where(~!all(is.na(.x)))) # Drop columns that contain all NA values
```

> This section uses the dataset published by Our World in Data, and gives an overview of Covid and vaccination trends in Canada.

### Country level trend

```{r}
cvv <- ggplot(CA, aes(x = date)) +
  geom_point(aes(y = new_cases, color = "Cases"), size = 0.75, shape = 21) +
  geom_point(aes(y = new_vaccinations, color = "Vaccinations"), size = 0.75, shape = 21) +
  geom_line(aes(y = new_cases_smoothed, color = "Cases(trend)"), size = 1) +
  geom_line(aes(y = new_vaccinations_smoothed, color = "Vaccinations(trend)"), size = 1) +
  theme_classic() +
  scale_color_manual(values = c("Cases(trend)" = "darkblue", "Vaccinations(trend)" = "palegreen4",
                                "Cases" = "skyblue1", "Vaccinations" = "darkseagreen")) +
  guides(color = guide_legend(override.aes = list(linetype = c("Cases(trend)" = 1,
                                                               "Vaccinations(trend)" = 1,
                                                               "Cases" = 0,
                                                               "Vaccinations" = 0),
                                                  shape = c("Cases(trend)" = NA,
                                                            "Vaccinations(trend)" = NA,
                                                            "Cases" = 21,
                                                            "Vaccinations" = 21)))) +
  labs(title = "Covid-19 daily cases and vaccinations in Canada",
       x = "Date",
       y = "Daily cases & vaccinations (log10 scale)",
       color = NULL, caption = "Data: OWID") +
  scale_x_date(date_breaks = "4 month") +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom") +
  scale_y_continuous(trans="log10")
ggplotly(cvv) %>%
  layout(hovermode = "x", legend = list(orientation = "h", xanchor = "center", x = 0.5, y = -0.1))
```

The log10 scale stabilizes the variance in the trend lines. It also allows a much easier comparison between the two trend lines since the vaccination number is much larger in scale than the case counts.

### Vaccination progress

```{r}
# Find out the rows where the second dose data are NAs when first dose were given at the start of the vaccination process
rows <- which(is.na(CA$people_fully_vaccinated) & !is.na(CA$people_vaccinated))
CA[rows, "people_fully_vaccinated"] = 0  # Change the NA values to 0

vaccine <- CA %>%
  mutate(`First dose` = round(people_vaccinated / population * 100, 3),
         `Fully vaccinated` = round(people_fully_vaccinated / population  * 100, 3),
          Booster = round(total_boosters / population * 100, 3)) %>%
  select(date, `First dose`, `Fully vaccinated`, Booster) %>%
  drop_na(`First dose`, Booster) %>%
  gather("doses", "Population share", -c(date)) %>%
  mutate(doses = factor(doses, levels = c("First dose", "Fully vaccinated", "Booster")))

vac <- ggplot(data = vaccine, aes(x = date, y = `Population share`, fill = doses)) +
  geom_area(position = "identity") +
  theme_classic() +
  scale_fill_brewer(palette = "Greens") +
  scale_x_date(date_breaks = "2 month") +
  labs(title = "Covid-19 vaccination progress in Canada ",
       x = "Date", y = "Share of Population(%)",
       fill = NULL, caption = "Data: OWID") +
  theme(plot.title = element_text(hjust = -0.25), legend.position = "bottom")
# vac +
#   transition_reveal(date) +
#   ease_aes()
ggplotly(vac) %>%
   layout(hovermode = "x", legend = list(orientation = "h", xanchor = "center", x = 0.5, y = -0.1))
```

```{r}
# Import JHU data
confirmedraw <- read_csv("https://github.com/CSSEGISandData/COVID-19/raw/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv") %>%
  filter(`Country/Region` == "Canada") %>%
  rename(Province = `Province/State`) %>%
  select(!`Country/Region`)
deathsraw <- read_csv("https://github.com/CSSEGISandData/COVID-19/raw/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv") %>%
  filter(`Country/Region` == "Canada") %>% 
  rename(Province = `Province/State`) %>%
  select(!`Country/Region`)

# Data cleaning: To create provincial level data in Canada
# Convert each data set from wide to long AND aggregate at provincial level  
confirmed <- confirmedraw %>%
  gather(key="date", value="confirmed", -c(Province, Lat, Long)) %>%
  mutate(date = mdy(date)) 
deaths <- deathsraw %>%
  gather(key="date", value="deaths", -c(Province, Lat, Long)) %>%
  mutate(date = mdy(date)) 

# Final data: combine both confirmed and deaths
CA_provinces <- left_join(confirmed, deaths, by = c("Province", "date", "Lat", "Long"))

# Create new variable: daily reported numbers of cases, deaths and recovered cases, 
#                      7 days rolling average case numbers, number of days, weeks
# Drop observations of Repatriated Travellers and the two cruise ships from the provincial data
CA_provinces <- CA_provinces %>% 
  filter(Province != "Repatriated Travellers" & Province != "Diamond Princess" &
         Province != "Grand Princess") %>% 
  group_by(Province) %>% 
  mutate(daily.confirmed = c(confirmed[1], diff(confirmed)),
         rolling.confirmed = round(rollmean(daily.confirmed, k = 7, fill = NA, align = "right"), 3),
         daily.deaths = c(deaths[1], diff(deaths)),
         rolling.deaths = round(rollmean(daily.deaths, k = 7, fill = NA, align = "right"), 3),
         days = date - first(date) + 1, 
         weeks = floor(as.numeric(difftime(date, first(date), units = "weeks"))) + 1)
```

> Since Our World in Data does not include provincial level data in its dataset, the following section uses the dataset published by Johns Hopkins University, and gives an overview of Covid and its related death count trends in Canadian provinces.

### Provincial level trend

Vaccination data isn't reported in the JHU dataset, the graph displays the death count trend along with the case trend instead.

```{r}
facet_case <- CA_provinces %>%
                drop_na(rolling.confirmed) %>%
                ggplot(aes(x = date)) + 
                    geom_point(aes(y = daily.confirmed, color = "Cases"), 
                               size = 0.25, shape = 21) +
                    geom_line(aes(y = rolling.confirmed, color = "Cases(trend)"), size = 0.5) +
                    geom_line(aes(y = rolling.deaths, color = "Deaths(trend)"), size = 0.5) +             
                    geom_bar(aes(y = daily.deaths), stat = "identity", width = 0.1, fill = "red") +
                    theme_bw() +
                    scale_color_manual(values = c("Cases(trend)" = "darkblue", 
                                                  "Cases" = "skyblue1",
                                                  "Deaths(trend)" = "coral2")) +
                    guides(color = guide_legend(override.aes = list(linetype = c("Cases(trend)" = 1, 
                                                                                 "Cases" = 0,
                                                                                 "Deaths(trend" = 1),
                                                                    shape = c("Cases(trend)" = NA, 
                                                                              "Cases" = 21,
                                                                              "Deaths(trend)" = NA)))) +
                    scale_y_continuous(trans="log10") +
                    scale_x_date(date_labels = "%b %Y", date_breaks = "4 months") +
                    labs(title = "COVID-19 daily case and death trends (log10 scale) in Canadian provinces",
                         x = "Date",
                         y = NULL,
                        # y = "Daily cases & death counts (log scale)",
                         color = NULL, fill = NULL, caption = "Data: JHU") +
                    theme(plot.title = element_text(hjust = 0.5), 
                          legend.position = "bottom",
                          axis.text.x = element_text(angle = 60, hjust = 1)) +
                    facet_wrap(facets = "Province")
ggplotly(facet_case) %>%
  layout(legend = list(orientation = "h", xanchor = "center", x = 0.5, y = -0.1))
```

### Total cases map

```{r}
map_ca <- ne_states(country = "Canada", returnclass = "sf") %>%
  rename(Province = gn_name) %>%
  select(Province, postal, region, geometry) %>%
  left_join(CA_provinces, by = "Province") %>%
  filter(date == max(date)) %>%
  rename(`Total cases` = confirmed) %>%
  ggplot() +
    geom_sf(aes(fill = `Total cases`, text = paste("Province: ", Province)),
            color = "gray40") +
    geom_sf_text(aes(label = postal), size = 2) +
    scale_fill_gradient("Total COVID cases", labels = scales::number, 
                        low = "#fee8c8", high = "#e34a33") +
    theme_void() +
    theme(panel.grid = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          plot.title = element_text(hjust=0.5,color="gray40", size=14),
          plot.subtitle = element_text(color="gray40", size = 10),
          legend.title	= element_text(color="grey35", size=9),
          legend.text	= element_text(color="grey35", size=9)) +
    labs(title = "Total Covid cases in Canadian provinces",
         subtitle = paste("As of", max(CA_provinces$date), '%d %B, %Y'))
ggplotly(map_ca, tooptip = "text") %>%
  layout(title = list(text = paste("Total Covid-19 cases in Canadian provinces<br>",
                                   "<sup>As of", max(CA_provinces$date), '</sup>')))
```

```{r total death map}
# map_ca_d <- ne_states(country = "Canada", returnclass = "sf") %>%
#   rename(Province = gn_name) %>%
#   select(Province, postal, region, geometry) %>%
#   left_join(CA_provinces, by = "Province") %>%
#   filter(date == max(date)) %>%
#   rename(`Total death counts` = deaths) %>%
#   ggplot() +
#     geom_sf(aes(fill = `Total death counts`, text = paste("Province: ", Province)),
#             color = "gray40") +
#     geom_sf_text(aes(label = postal), size = 2) +
#     scale_fill_gradient("Total death counts", labels = scales::number, 
#                         low = "#fee8c8", high = "#e34a33") +
#     theme_void() +
#     theme(panel.grid = element_blank(),
#           axis.text = element_blank(),
#           axis.ticks = element_blank(),
#           plot.title = element_text(hjust=0.5,color="gray40", size=14),
#           plot.subtitle = element_text(color="gray40", size = 10),
#           legend.title	= element_text(color="grey35", size=9),
#           legend.text	= element_text(color="grey35", size=9)) +
#     labs(title = "Total Covid death counts in Canadian provinces",
#          subtitle = paste("As of", max(CA_provinces$date), '%d %B, %Y'))
# ggplotly(map_ca_d, tooptip = "text") %>%
#   layout(title = list(text = paste("Total Covid-19 deaths in Canadian provinces<br>",
#                                    "<sup>As of", max(CA_provinces$date), '</sup>')))
```

### New cases map for the past 6 weeks

Since some provinces do not report Covid cases during the weekend, displaying the weekly data would be more useful than the daily result.

```{r}
ca <- CA_provinces %>%
  filter(weeks >= max(weeks) - 5) %>%
  group_by(weeks, Province) %>%
  summarize(daily.confirmed = sum(daily.confirmed),
            daily.deaths = sum(daily.deaths),
            label = min(date)) %>%
              # paste0(min(format(date, "%y/%m/%d")), "-", max(format(date, "%y/%m/%d")))) %>%
  mutate(# label = gsub(label, pattern = "21/|22/", replacement = ""),
         text = paste("Province:", Province, "\nConfirmed cases:", daily.confirmed),
         text2 = paste("Province:", Province, "\nDeaths:", daily.deaths))
  
ne_states(country = "Canada", returnclass = "sf") %>%
  rename(Province = gn_name) %>%
  select(Province, geometry) %>%
  left_join(ca, by = "Province") %>%
  plot_ly(stroke = I("black"), split = ~Province, color = ~daily.confirmed, 
          colors = viridis::inferno(99),
          text = ~text, showlegend = FALSE, hoveron = "fills", frame = ~label) %>%
            # gsub(label, pattern = "21/|22/", replacement = "")) %>%
  colorbar(title = "Weekly confirmed cases") %>%
  style(hoverlabel = list(bgcolor = "white")) %>%
  animation_slider(currentvalue = list(prefix="Date: ", font = list(color = "red"))) %>%
  layout(title = list(text = paste("New weekly Covid-19 cases in Canadian provinces<br>",
                                   "<sup>As of", max(CA_provinces$date), '</sup>')))
```

```{r New death maps for the past 6 weeks}
# ne_states(country = "Canada", returnclass = "sf") %>%
#   rename(Province = gn_name) %>%
#   select(Province, geometry) %>%
#   left_join(ca, by = "Province") %>%
#   plot_ly(stroke = I("black"), split = ~Province, color = ~daily.deaths, colors = viridis::inferno(99),
#           text = ~text2, showlegend = FALSE, hoveron = "fills", frame = ~label) %>%
#   colorbar(title = "Weekly death toll") %>%
#   style(hoverlabel = list(bgcolor = "white")) %>%
#   animation_slider(currentvalue = list(prefix="Date: ", font = list(color = "red"))) %>%
#   layout(title = list(text = paste("New weekly Covid-19 deaths in Canadian provinces<br>",
#                                    "<sup>As of", max(CA_provinces$date), '</sup>')))
```
